Legacy Data Confounds Genomics Studies

### 2019-08-08 LatexManuscript/MBE/legacy.tex
moved things to one folder for mbe upload

\maketitle
\begin{abstract}
Recent reports have identified differences in the mutational spectra across human populations. While some of these reports have been replicated in other cohorts, most have been reported only in the 1000 Genomes Project (1kGP) data. While investigating an intriguing putative population stratification within the Japanese population, we identified a batch effect leading to spurious mutation calls in the 1kGP data and to the apparent population stratification. Because the 1kGP data is used extensively, we find that the batch effects also lead to incorrect imputation by leading imputation servers and a small number of suspicious GWAS associations. Lower-quality data from the early phases of the 1kGP thus continues to contaminate modern studies in hidden ways. It may be time to retire or upgrade such legacy sequencing data.
\vspace{5mm}

@@

The last 5 years have seen a drastic increase in the amount and quality of human genome sequence data. 
Reference cohorts such as the International HapMap Project \citep{HapMap2005}, the 1000 Genomes Project (1kGP)\citep{1000GenomesProjectConsortium2010,The1000GenomesProjectConsortium2012,phase3}, and the Simons Diversity project \citep{Mallick2016}, for example, have made thousands of genome sequences publicly available for population and medical genetic analyses. 
Many more genomes are available indirectly through servers providing imputation services \citep{mccarthy2016reference} or summary statistics for variant frequency estimation \citep{Lek2016}.

@@

Yet, because of the extraordinary value of freely available data, early data from the 1kGP is still widely used to impute untyped variants to estimate allele frequencies and to answer a wide range of medical and evolutionary questions.
This raises the question of whether and how such legacy data should be included in contemporary analyses alongside more recent cohorts.
Mafessoni et al. recently identified batch effects in the 1kGP by looking for individuals with excess LD among distant variants. \citep{mafessoni2018turning}.
Here we point out how these and additional unreported batch effects in the early phases of the 1kGP lead to incorrect genetic conclusions through population genetic analyses and spurious GWAS associations as a result of imputation using the 1kGP as a reference.

@@

In particular, the latter two studies  identified a heterogeneous mutational signature within 1kGP Japanese individuals.
This heterogeneity is intriguing because differences in germline signatures accumulate over many generations.
A systematic difference within the Japanese population would suggest sustained environmental or genetic differences across sub-populations within Japan with little to no gene flow.
Given that this signature had not yet been reproduced in another dataset, we decided to follow up on this observation, by using a newly sequenced dataset of Japanese individuals from Nagahama.

While we were unable to reproduce the mutational heterogeneity within the Japanese population, we could trace back the source of the discrepancy to a technical artefact in the 1kGP data.
In addition to creating biases in mutational signatures, this artefact leads to spurious imputation results which have found their way in recent publications and online resources.

@@

\begin{figure}
\scalebox{0.85}{\includegraphics[width=\hsize,keepaspectratio]{../Figures/Figure1.jpg}}
\caption{
Suspicious mutations carried by individuals with low quality data have distinct mutational profiles, reproduce poorly across studies, and are distributed across the genome.
\textbf{A} 
Genome wide association of the average quality of mapped bases $Q$ for the 104 Japanese individuals included in the 1000 Genomes Project. 
This GWAS identified $1034$ SNPs associated to the average $Q$ of SNPs mapped for an individual, with  $p < 10^{-6}$
($587$ SNPs had  $p < 10^{-8}.$)
\textbf{B} 
Joint frequency spectrum plot of the Japanese from the 1000 Genomes Project and a more recent Japanese dataset from Nagahama.
The plot is zoomed in on frequencies below 0.5 for clarity.
The size of blue dots are proportional to the number of suspicious SNPs in a given frequency bin (out of the 1034 identified by GWAS).  
\textbf{C} 
Mutation spectrum of these 1034 variants compared to the expectation from the distribution of all SNPs.
The majority of the variants with significant associations to $Q$ have the *AC${\rightarrow}$*CC mutational pattern. 
There is also an enrichment in GA*${\rightarrow}$GG* and GC*${\rightarrow}$GG* mutations. 
These three enrichments can be summarized as G**${\rightarrow}$GG*. 
Stars ( * ) indicate a significant deviation from the expected mutational spectrum from all SNPs.
}
 \label{SFS}
\end{figure}

@@

Harris and Pritchard reported an excess of a 3-mer substitution patterns *AC${\rightarrow}$*CC in a portion of the Japanese individuals in the 1kGP \citep{Harris2017a}.
Our initial goal was to determine whether this signature could be explained by population structure or by technological bias.
While trying to follow up on this observation in a larger and more recent Japanese cohort from Nagahama, we did not find this particular signature.
When comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed a number of single nucleotide polymorphisms (SNPs) private to one of the two groups (Figure \ref{SFS}B).
Given the similarity of the two populations, this strongly suggests a technical difference rather than a population structure effect.
These mismatches were maintained despite only considering sites that satisfied strict quality masks and Hardy-Weinberg equilibrium in both cohorts.

@@

\begin{figure}
\includegraphics[width=0.95\hsize,keepaspectratio]{../Figures/MapQualOverTime.jpg}

@@

To identify SNPs that are likely to reproduce poorly across cohorts without having access to a second cohort, we performed an association study in the JPT for SNPs that associate strongly with low $Q$ (Figure \ref{SFS}).
Traditionally, genome wide association studies use genotypes as the independent variable. 
Here we perform a genotype conditional association test (GCAT), where genotypes are now the dependent variable that we predict using the continuous variable $Q$ as the independent variable \citep{song2015testing}.
We use logistic regression of the genotypes on $Q$ and identify 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$. 
While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., excluding SNPs with $ p < 10^{-6}$ is more stringent than excluding SNPS with $p < 10^{-8}$). 
The variants that are associated to $Q$ have a significant enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations (Figure \ref{SFS}C).
These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in individuals with low $Q$.
Statistical significance of these enrichments is computed using a chi-squared test following \cite{Harris2017a}.

Thus, this mutational signal is heavily enriched in $Q$-associated SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low $Q$ remain unidentifiable using association techniques due to lack of power (Supplementary Figure \ref{histogram_of_sig_snps}).
The removal of individuals with $Q$ below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged (Figure \ref{MutSpect}).
For population genetic analyses sensitive to the accumulation of rare variants, the removal of individuals with low $Q$ appears preferable to filtering specific low-quality SNPs. 
For other analyses where quality of imputation matters, identifying and discarding $Q$-associated variants may be preferable.

@@

We therefore performed the same association approach in all populations independently, and similarly identified $Q$-associated SNPs in 23 of the 26 populations in the 1kGP, with the phase 1 populations being most affected, with on average four times as many significantly associated sites compared to the phase 3 populations. Over 812 variants were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ in each.

To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with the logistic factor analysis (LFA) as an offset to account for population structure. This approach is called Genotype-Conditional Association Test  (GCAT) as proposed by \citep{song2015testing}. 
We also considered two alternative approaches to account for confounders, namely using the leading five  principal components, and using population membership as covariates. 
These models were broadly consistent (See Figure \ref{CompareModel}).

The GCAT method identifies a total of 24,390 variants associated to $Q$ distributed across the genome with 15,270 passing the 1kGP strict mask filter (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}). Most analyses below focus on the 15,270 variants satisfying the strict mask, since these variants are unlikely to be filtered by standard pipelines.
To account for the large number of tests, we used a two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the p-values using a nominal Type-I error rate $\alpha = 0.01$ \citep{Benjamini2006}. 
We tested SNPs, INDELs and repetitive regions separately as they may have different error rates (Table \ref{sigTable}).
Lists of $Q$-associated variants and individuals with low $Q$ are provided in Supplementary Data.

@@

A recent publication by Mafessoni et al. also identified a batch effect in the 1kGP using a method that uses linkage disequilibrium rather than quality metrics to identify 19,196 suspicious variants with 67\% of them passing the 1kGP strict mask \citep{mafessoni2018turning} (Figure \ref{Mafessoni_MutSpect}A).
They identify 17,917 variants significantly associated to abnormal LD patterns that are not associated to $Q$.
We find that 1,279 ( 3\% ) of the variants they identified are also in our list of suspicious variants and have correlated p-values to those identified using the GCAT method.
We also find 23,111 $Q$-associated variants that are not associated to abnormal LD patterns. 
Interestingly, the variants identified by Mafessoni et al. are not enriched in the mutational spectra described above (Figure \ref{Mafessoni_MutSpect}B).
These results indicate that there may be a multitude of batch effects in the 1kGP that can only be identified using a suite of association tests.

\begin{table}[h!]
\centering
\begin{tabular}{l  r r r  r r}
                      & {Repeat} & \textcolor{gray}{Proportion} & {Non-Repeat} & \textcolor{gray}{Proportion}   & {Total}    \\ \hline
{SNP}  & 3,369 &\textcolor{gray}{  $5.3\times 10^{-4}$}& 11,059 &\textcolor{gray}{  $5.6\times 10^{-4}$}& 14,428\\  
{INDEL} & 181& \textcolor{gray}{  $3.0\times 10^{-4}$}& 657& \textcolor{gray}{  $6.6\times 10^{-4}$} & 838\\ \hline
{Total} & 3,550& & 11,716 & & 15,270\\
\end{tabular}
\caption{Number and proportion of $Q$-associated variants passing the 1000 Genomes Project strict mask per category. 
Variants that are flagged by the 1000 Genomes Project nested repeat mask file were analyzed separately for FDR calculation. SNPs and INDELs were also analyzed separately.
A total of 15,270 are statistically significantly associated to $Q$ and pass the 1kGP strict mask. 
The grey text is the proportion of $Q$-associated variants per category.
The number of variants included in the analysis for SNPs, SNPs in repeat regions, INDELs and INDELs in repeat regions are  19,846,786, 6,312,620, 1,770,315 and 586,342 respectively.}
\label{sigTable}
\end{table}

@@

We did a similar analysis using all variants identified in the GCAT model (rather than only variants significantly associated to $Q$ within the CHB and CHS). 
Of the 15,270 $Q$-associated variants identified globally, 6,307 are polymorphic in the 1kGP for the 83 resequenced individuals  (See Figure \ref{90HanSFS_full}).
The vast majority of polymorphisms associated with $Q$ are not present at all in the resequencing dataset, suggesting that they are spurious variants.
Five variants show differing frequencies between both datasets and are likely explained by biased genotypes.
Finally, 1,139 (or 18\%) are present in the resequenced data at comparable frequencies.
Because this is higher than the false discovery rate, we conclude that the cohort-wide GCAT test identifies predominantly spurious variants, but also variants that have biased genotypes. The similarity in allele frequencies for these variants suggests that the biases in allele frequencies are present in populations other than the CHB and CHD.

@@

To investigate whether suspicious calls from the 1kGP are imputed into genotyping studies, we submitted genotype data for the first two chromosomes of the 1kGP genotype data to the Michigan Imputation Server.
We found that all of the variants associated with $Q$ were imputed back in the samples.
This suggests that the imputation reference panel still includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the individuals with low $Q$.
These $Q$-associated variants could also compromise the imputation of nearby real variants because haplotype matching would be compromised. However when considering the imputation scores of genotype data from Japanese individuals from Nagahama, there does not appear to be any impact on nearby variant imputation scores (Figure \ref{imputation_score}).

We searched the literature for GWAS that have reported these $Q$-associated variants as being significantly associated with some biological trait, even though there is no particular reason for these variants to be associated with phenotypes.
The NHGRI-EBI Catalog of published genome-wide association studies identified seventeen recent publications that had reported these variants as close to or above the genome-wide significant threshold (Table \ref{gwasTable}).

@@

These articles used a variety of strict quality filters, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
They also removed rare alleles and alleles with high degrees of missingness.
%Indeed, we expect a large number of $Q$-associated variants to be filtered out by some quality controls like the Hardy-Weinberg equilibrium test.
Despite state-of-the-art quality controls, the variants were imputed onto genotype data, reached genome wide significance for association with biological traits, and were reported to the GWAS catalogue.
Removing these variants from imputation servers could help avoid spurious associations.

These associations are not necessarily incorrect -- a weak but significant bias in imputation may still result in a correct association.
To distinguish between variants with weak but significant association with $Q$ from variants with strong biases, we distinguished between variants where the allele frequency difference between individuals with low- and high-$Q$ is larger than a factor of two (which naturally separates two clusters of variants on Figure \ref{90HanSFS_full}).
The majority (92.7\%) of the $Q$-associated variants are strongly biased in that they are more than twice as frequent in individuals with low-$Q$ compared to high-$Q$ data.
By contrast, most $Q$-associated variants reported in the GWAS catalogue had weak bias (See Figure \ref{Pub_30}), with three exceptions.
One study that reports associations with seven highly biased $Q$-associated variants considered copy number of Epstein-Barr virus (EBV) sequence in the 1kGP as a phenotype \citep{Mandage2017}. 
Even though these seven variants are reported in the GWAS catalogue, they were not flanked by LD peaks and were correctly removed from downstream analyses by \citep{Mandage2017}. 
It is plausible that the EBV copy number phenotype estimation is sensitive to mapping and confounded by the same technical artefacts that lead to biased SNP calling.

@@

To pinpoint the precise technical source of the discrepancy would require further forensic inquiries into the details of the heterogeneous sample preparation and data processing pipelines used throughout the 1kGP. 
Given the progress in sequencing and calling that occurred since the early phases of the 1kGP (Figure \ref{MapQual}), it is likely that the source of these biases is not longer actively introducing bias in recent sequence data.

@@

On a technical front, we were surprised that such strong associations between variants and technical covariates in the 1kGP project had not been identified before. 
The genome-wide logistic regression analysis of genotype on quality metric is straightforward, and should probably be a standard in a variety of -omics studies. 
The logistic factor analysis is more computationally demanding but produces more robust results \citep{song2015testing}.

@@

\section{Methods}
\subsection{Code and data availability}
Since this analysis is primarily performed using publicly available data, we provide fully reproducible and publicly available pipelines on \href{https://github.com/LukeAndersonTrocme/LegacyData}{GitHub}.
This repository includes scripts used for data download, processing, analysis and plotting.
%Supplementary data as well as preprocessed data for plotting are also available on GitHub.

@@

\subsection{Quality Controls}
For the mutation spectrum analysis, we reproduced the quality control and data filtering pipelines used by Harris et al. 
Several mask files were applied to remove regions of the genome that might be lower quality, or might have very different mutation rates or base pair complexity compared to the rest of the genome. 
The 1kGP strict mask was used to remove low quality regions of the genome, highly conserved regions were removed using the phastCons100way mask file and highly repetitive regions were removed using the NestedRepeats mask file from RepeatMasker.
Furthermore, only sites with missingness below 0.01, MAF less than 0.1, and MAF greater than 0.9 were considered.
In total, 7,786,023 diallelic autosomal variants passed our quality controls for the mutation spectrum analysis.
We calculated the mutation spectrum of base pair triplets for the list of significant variants for the JPT population using a similar method as described in  \citep{Harris2017a}.

For the GCAT analysis, the only filtration used was the application of an minor and major allele frequency cutoff of 0.000599 (removing singletons, doubletons and tripletons) resulting in a total of S=28,516,063 variants included in the test. 
We also used the NestedRepeats mask file to flag variants inside repetitive regions as these were analyzed separately for false discovery rate estimation.
Variants flagged by the 1kGP strict mask are included in the association test and included in the four separate FDR adjustments for the four categories of variants (INDELs in Repeat and Non-Repeat regions and SNPs in Repeat and Non-Repeat regions). 
These variants are only removed after the FDR and excluded from downstream discussion of error patterns, since most population genetics analyses use the strict mask as a filter, and we expect to find problematic variants in filtered regions.

@@

Under the null hypothesis the slope coefficient $\beta_s$ is zero and Model \eqref{gcat_basic} reduces to Model \eqref{lfa_basic}. 
$\beta_s$ denotes the association to average quality of mapped bases $Q$ to genotype $y_{s}$. 
To test the null hypothesis, we use the generalized likelihood ratio test statistic, whose deviance is a measure of the marginal importance of adding $Q$ in the model. 
The deviance test statistic under the null model is approximately chi-square distributed with one degree of freedom.

We run a total of $S$ regressions, where $S$ is the total number of genomic loci. Given the large number of tests, the large proportion of expected null hypotheses and the positive dependencies across the genome, we used the two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the \textit{p}-values \citep{Benjamini2006}.
By using a nominal Type-I error rate $\alpha = 0.01$, a total of 15,270 variants were found to be statistically significant. 
See Supplementary Data for a list of variants and adjusted \textit{p}-values.

@@

\begin{figure}
\centering
\includegraphics[width=13cm,keepaspectratio]{../Figures/fits_Significant_Positions_CompareModels.jpg}
\caption{Comparison of three logistic regression models for testing association to $Q$.
These methods model each individual-specific allele frequency as a logistic function using principal components (PC), Population membership (Pop) or LFA as an offset.
In these plots we are comparing the deviance from the null model in the 15,270 variants identified using the LFA model.}  
\label{CompareModel}
\end{figure}

>>>>

\maketitle
\begin{abstract}
Recent reports have identified differences in the mutational spectra across human populations. While some of these reports have been replicated in other cohorts, most have been reported only in the 1000 Genomes Project (1kGP) data. While investigating an intriguing putative population stratification within the Japanese population, we identified a previously unreported batch effect leading to spurious mutation calls in the 1kGP data and to the apparent population stratification. Because the 1kGP data is used extensively, we find that the batch effects also lead to incorrect imputation by leading imputation servers and a small number of suspicious GWAS associations. Lower-quality data from the early phases of the 1kGP thus continues to contaminate modern studies in hidden ways. It may be time to retire or upgrade such legacy sequencing data.
\vspace{5mm}

@@

The last 5 years have seen a drastic increase in the amount and quality of human genome sequence data. 
Reference cohorts such as the International HapMap Project \citep{HapMap2005}, the 1000 Genomes Project (1kGP)\citep{1000GenomesProjectConsortium2010,The1000GenomesProjectConsortium2012,phase3}, and the Simons Genome Diversity project (SGDP)\citep{Mallick2016}, for example, have made thousands of genome sequences publicly available for population and medical genetic analyses. 
Many more genomes are available indirectly through servers providing imputation services \citep{mccarthy2016reference} or summary statistics for variant frequency estimation \citep{Lek2016}.

@@

Yet, because of the extraordinary value of freely available data, early data from the 1kGP is still widely used to impute untyped variants, to estimate allele frequencies, and to answer a wide range of medical and evolutionary questions.
This raises the question of whether and how such legacy data should be included in contemporary analyses alongside more recent cohorts.
Mafessoni et al. recently identified batch effects in the 1kGP by looking for individuals with excess LD among distant variants. \citep{mafessoni2018turning}.
Here we point out how these and additional unreported batch effects in the early phases of the 1kGP lead to incorrect genetic conclusions through population genetic analyses and spurious GWAS associations as a result of imputation using the 1kGP as a reference.

@@

In particular, the latter two studies  identified a heterogeneous mutational signature within 1kGP Japanese individuals.
This heterogeneity is intriguing because differences in germline signatures accumulate over many generations.
A systematic difference within the Japanese population would suggest sustained environmental or genetic differences across sub-populations within Japan with little to no gene flow.
\todo{While this signature had been partially reproduced in the SGDP, it had yet to be reproduced in a Japanese dataset, we decided to follow up on this observation, by using a newly sequenced dataset of Japanese individuals from Nagahama \citep{Harris2017a,Aikens2018}.}
While we were unable to reproduce the mutational heterogeneity within the Japanese population, we could trace back the source of the discrepancy to a technical artefact in the 1kGP data.
In addition to creating biases in mutational signatures, this artefact leads to spurious imputation results which have found their way in recent publications and online resources.

@@

\begin{figure}
\scalebox{0.85}{\includegraphics[width=\hsize,keepaspectratio]{Figure1.jpg}}
\caption{
Suspicious mutations carried by individuals with low quality data have distinct mutational profiles, reproduce poorly across studies, and are distributed across the genome.
\textbf{A} 
Genome wide association of the average quality of mapped bases $Q$ for the 104 Japanese individuals included in the 1000 Genomes Project. 
This GWAS identified $1034$ SNPs associated to the average $Q$ of SNPs mapped for an individual, with  $p < 10^{-6}$
($587$ SNPs had  $p < 10^{-8}.$)
\textbf{B} 
Joint frequency spectrum plot of the Japanese from the 1000 Genomes Project and a more recent Japanese dataset from Nagahama.
The plot is zoomed in on frequencies below 0.5 for clarity.
The size of blue dots are proportional to the number of variants in a given frequency bin that associate with $Q$ in the JPT. 
\textbf{C} 
Mutation spectrum of the 1034 variants that associated with $Q$ in the JPT($p < 10^{-6}$), compared to the expectation from the distribution of all SNPs.
The majority of the variants with significant associations to $Q$ have the *AC${\rightarrow}$*CC mutational pattern. 
There is also an enrichment in GA*${\rightarrow}$GG* and GC*${\rightarrow}$GG* mutations. 
These three enrichments can be summarized as G**${\rightarrow}$GG*. 
Stars ( * ) indicate a significant deviation from the expected mutational spectrum from all SNPs.
}
 \label{SFS}
\end{figure}

@@

Harris and Pritchard reported an excess of a 3-mer substitution patterns *AC${\rightarrow}$*CC in a portion of the Japanese individuals in the 1kGP \citep{Harris2017a}.
Our initial goal was to determine whether this signature could be explained by population structure or technological error.
While trying to follow up on this observation in a larger and more recent Japanese cohort from Nagahama, we did not find this particular signature.
When comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed a number of single nucleotide polymorphisms (SNPs) private to one of the two groups (Figure \ref{SFS}).
Given the similarity of the two populations, this strongly suggests a technical difference rather than a population structure effect.
These mismatches were maintained despite only considering sites that satisfied strict quality masks and Hardy-Weinberg equilibrium in both cohorts.

@@

\begin{figure}
\includegraphics[width=0.95\hsize,keepaspectratio]{MapQualOverTime.jpg}

@@

To identify SNPs that are likely to reproduce poorly across cohorts without having access to a second cohort, we performed an association study in the JPT for SNPs that associate strongly with low $Q$ (Figure \ref{SFS}).
Traditionally, genome wide association studies use genotypes as the independent variable. 
Here we perform a genotype conditional association test (GCAT), where genotypes are now the dependent variable that we predict using the continuous variable $Q$ as the independent variable \citep{song2015testing}.
We use logistic regression of the genotypes on $Q$ and identify 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$. 
While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., excluding SNPs with $ p < 10^{-6}$ is more stringent than excluding SNPS with $p < 10^{-8}$). 
The variants that are associated to $Q$ have a significant enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations (Figure \ref{SFS}C).
These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in individuals with low $Q$.
Statistical significance of these enrichment is computed using a chi-squared test following \cite{Harris2017a}.

Thus, this mutational signal is heavily enriched in $Q$-associated SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low $Q$ remain unidentifiable using association techniques due to lack of power (Supplementary Figure \ref{histogram_of_sig_snps}).
The removal of individuals with $Q$ below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged (Figure \ref{MutSpect}).
For population genetic analyses sensitive to the accumulation of rare variants, the removal of individuals with low $Q$ appears preferable to filtering specific low-quality SNPs. 
For other analyses where quality of imputation matters, identifying $Q$-associated variants may be preferable.

@@

We therefore performed the same reverse GWAS approach in all populations independently, and similarly identified $Q$-associated SNPs in 23 of the 26 populations in the 1kGP, with the phase 1 populations being most affected, with on average four times as many significantly associated sites compared to the phase 3 populations.
Over 1,165 variants were independently associated to low $Q$ with $ p < 10^{-6}$ in each (Figure \ref{OverLap}).

To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with the logistic factor analysis (LFA) as an offset to account for population structure or Genotype-Conditional Association Test  (GCAT) as proposed by \citep{song2015testing}. 
We also considered two alternative approaches to account for confounders, namely using the leading five  principal components, and using population membership as covariates. 
These models were broadly consistent (See Figure \ref{CompareModel}).

This method identifies a total of 24,390 variants associated to $Q$ distributed across the genome with 15,270 passing the 1kGP strict mask filter (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}). Most analyses below focus on the 15,270 variants satisfying the strict mask, since these variants are unlikely to be filtered by standard pipelines.
To account for the large number of tests, we used a two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the p-values using a nominal Type-I error rate $\alpha = 0.01$ \citep{Benjamini2006}. 
We tested SNPs, INDELs and repetitive regions separately as they may have different error rates (Table \ref{sigTable}).
Lists of $Q$-associated variants and individuals with low $Q$ are provided in Supplementary Data.

@@

A recent publication by Mafessoni et al. also identified a batch effect in the 1kGP using a method that uses linkage disequilibrium rather than quality metrics to identify 19,196 suspicious variants with 67\% of them passing the 1kGP strict mask \citep{mafessoni2018turning} (Figure \ref{Mafessoni_MutSpect}A).
They identify 17,917 variants significantly associated to abnormal LD patterns that are not associated to $Q$.
We find that 1,279 ( 3\% ) of the variants they identified are also in our list of suspicious variants and have correlated p-values to those identified using the GCAT method.
We also find 23,111 $Q$-associated variants that are not associated to abnormal LD patterns. 
Interestingly, the variants identified my Mafessoni et al. are not enriched in the mutational spectra described above (Figure \ref{Mafessoni_MutSpect}B).
These results indicate that there may a multitude of batch effects in the 1kGP that can only be identified using a suite of association tests.

\begin{table}[h!]
\centering
\begin{tabular}{l  r r r  r r}
                      & {Repeat} & \textcolor{gray}{Proportion} & {Non-Repeat} & \textcolor{gray}{Proportion}   & {Total}    \\ \hline
{SNP}  & 3,369 &\textcolor{gray}{0.53\textperthousand} & 11,059 & \textcolor{gray}{0.56\textperthousand} & 14,428\\  
{INDEL} & 181& \textcolor{gray}{0.3\textperthousand} & 657& \textcolor{gray}{0.66\textperthousand} & 838\\ \hline
{Total} & 3,550& & 11,716 & & 15,270\\
\end{tabular}
\caption{Number and proportion of $Q$-associated variants passing the 1000 Genomes Project strict mask per category. 
Variants that are flagged by the 1000 Genomes Project nested repeat mask file were analyzed separately for FDR calculation. SNPs and INDELs were also analyzed separately.
A total of 15,270 are statistically significantly associated to $Q$ and pass the 1kGP strict mask. 
The grey text is the proportion of $Q$-associated variants per category.
The number of variants included in the analysis for SNPs, SNPs in repeat regions, INDELs and INDELs in repeat regions are  19,846,786, 6,312,620, 1,770,315 and 586,342 respectively.}
\label{sigTable}
\end{table}

@@

We did a similar analysis using all variants identified in the GCAT model (rather than only variants significantly associated to $Q$ within the CHB and CHS). 
Of the 15,270 $Q$-associated variants identified globally, 6,307 are polymorphic in the 1kGP for the 83 resequenced individuals  (See Figure \ref{90HanSFS_full}).
The vast majority of polymorphisms associated with $Q$ are not present at all in the resequencing dataset, suggesting that they are spurious variants.
Five variants show differing frequencies between both datasets and are likely explained by biased genotypes.
Finally, 1,139 (or 18\%) are present in the resequenced data at comparable frequencies.
Because this is higher than the false discovery rate, we conclude that the cohort-wide GCAT test identifies predominantly spurious variants, but also variants that have biased genotypes in populations other than the CHB/CHD.

@@

To investigate whether suspicious calls from the 1kGP are imputed into genotyping studies, we submitted genotype data for the first two chromosomes of the 1kGP genotype data to the Michigan Imputation Server.
We found that all of the variants associated with $Q$ were imputed back in the samples.
This suggests that the imputation reference panel still includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the low-quality individual.
These $Q$-associated variants could also compromise the imputation of nearby real-variants, however when considering the imputation scores of genotype data from Japanese individuals from Nagahama, there does not appear to be any impact on nearby variant imputation scores (Figure \ref{imputation_score}).

We searched the literature for any GWAS that might have reported these $Q$-associated variants as being significantly associated with some biological trait, even though there is no particular reason for these variants to be associated with phenotypes.
The NHGRI-EBI Catalog of published genome-wide association studies identified seventeen recent publications that had reported these variants as close to or above the genome-wide significant threshold (Table \ref{gwasTable}).

@@

All of these articles used a variety of strict quality filters, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
They also removed rare alleles and alleles with high degrees of missingness.
Indeed, we expect a large number of $Q$-associated variants to be filtered out by some quality controls like the Hardy-Weinberg equilibrium test.
Even though the studies used state-of-the-art quality controls, the variants were imputed onto genotype data and reached genome wide significance for association with biological traits.
However, the fact that some of these variants in other studies are not removed and that the great majority of these variants are missing from higher quality datasets means that these $Q$-associated variants should be flagged for removal to avoid spurious association.

These associations are not necessarily incorrect -- a weak but significant bias in imputation may still result in a correct associations.
To distinguish between variants with weak but significant association with $Q$ from variants with strong biases, we distinguished between variants where the allele frequency difference between individuals with low- and high-$Q$ is larger than a factor of two (which naturally separates two clusters of variants on Figure \ref{90HanSFS_full}).
The majority (92.7\%) of the $Q$-associated variants are strongly biased in that they are more than twice as frequent in individuals with low-$Q$ compared to high-$Q$ data.
By contrast, most $Q$-associated variants reported in the GWAS catalogue had weak bias (See Figure \ref{Pub_30}), with three exceptions.
One study that reports associations with seven highly biased $Q$-associated variants considered copy number of Epstein-Barr virus (EBV) sequence in the 1kGP as a phenotype \citep{Mandage2017}. 
These seven variants were not flanked by LD peaks and were correctly removed from downstream analyses however, they were still included in the NHGRI-EBI Catalog \citep{Mandage2017}.
It is plausible that the EBV copy number phenotype is sensitive to mapping and confounded by the same technical artefacts that lead to biased SNP calling.

@@

To pinpoint the precise technical source of the discrepancy would require further forensic inquiries into the details of the heterogeneous sample preparation and data processing pipelines used throughout the 1kGP. 
Given the progress in sequencing and calling that occurred since the early phases of the 1kGP (Figure \ref{MapQual}), it is likely that the source of these biases is not longer being actively introduced in recent sequence data.

@@

On a technical front, we were surprised that strong association between variants and technical covariates in the 1kGP project had not been identified before. 
The genome-wide logistic regression analysis of genotype on quality metric is straightforward, and should probably be a standard in a variety of -omics studies. 
The logistic factor analysis is more computationally demanding but produces more robust results \citep{song2015testing}. 
All three approaches accounting for population structure we tested produce comparable results.

@@

\section{Methods}
\subsection{Code and data availability}
Since this data is primarily performed using publicly available data, we provide fully reproducible and publicly available on \href{https://github.com/LukeAndersonTrocme/LegacyData}{GitHub}.
This repository includes scripts used for data download, processing, analysis and plotting.
%Supplementary data as well as preprocessed data for plotting are also available on GitHub.

@@

\subsection{Quality Controls}
For the mutation spectrum analysis, we reproduced the quality control and data filtering pipelines used by Harris et al. as they applied the current state of the art quality thresholds to remove questionable sequences for detecting population level differences. 
Several mask files were applied to remove regions of the genome that might be lower quality, or might have very different mutation rates or base pair complexity compared to the rest of the genome. 
The 1kGP strict mask was used to remove low quality regions of the genome, highly conserved regions were removed using the phastCons100way mask file and highly repetitive regions were removed using the NestedRepeats mask file from RepeatMasker.
Furthermore, only sites with missingness below 0.01, MAF less than 0.1, and MAF greater than 0.9 were considered.
In total, 7,786,023 diallelic autosomal variants passed our quality controls for the mutation spectrum analysis.
We calculated the mutation spectrum of base pair triplets for the list of significant variants for the JPT population using a similar method as described in  \citep{Harris2017a}.

For the reverse GWAS, the only filtration used was the application of an minor and major allele frequency cutoff of 0.000599 (removing singletons, doubletons and tripletons) resulting in a total of S=28,516,063 variants included in the test. 
We also used the NestedRepeats mask file to flag variants inside repetitive regions as these were analyzed separately for false discovery rate estimation.
Variants flagged by the 1kGP strict mask are included in the association test and included in the FDR adjustment. 
These variants are only removed after the FDR and excluded from downstream discussion of error patterns, since most population genetics analyses use the strict mask as a filter, and we expect to find problematic variants in filtered regions.

@@

Under the null hypothesis the slope coefficient $\beta_s$ is zero and Model \eqref{gcat_basic} reduces to Model \eqref{lfa_basic}. 
$\beta_s$ denotes the association to average quality of mapped bases $Q$ to genotype $y_{s}$. 
To test the null hypothesis, we use the generalized likelihood ratio test statistic, whose deviance is a measure of the marginal importance of adding $Q$ in the model. 
The deviance test statistic under the null model is approximately chi-square distributed with one degrees of freedom.

We run a total of $S$ regressions, where $S$ is the total number of genomic loci. Given the large number of tests, the large proportion of expected null hypotheses and the positive dependencies across the genome, we used the two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the \textit{p}-values \citep{Benjamini2006}.
By using a nominal Type-I error rate $\alpha = 0.01$, a total of 15,270 variants were found to be statistically significance. 
See Supplementary Data for a list of variants and adjusted \textit{p}-values.

@@

\begin{figure}
\centering
\includegraphics[width=13cm,keepaspectratio]{fits_Significant_Positions_CompareModels.jpg}
\caption{Comparison of three logistic regression models for testing association to $Q$.
These methods model each genotype as a logistic function using principal components (PC), Population membership (Pop) or LFA as an offset.
In these plots we are comparing the deviance from the null model in the 15,270 variants identified using the LFA model.}  
\label{CompareModel}
\end{figure}
